Deep learning for quality control of surface physiographic fields using satellite Earth observations

### 2023-05-21 docs/submission/template.tex
more plots

\noindent As focus of this study are lake related fields, and lakes occupy only 1.8\% of the Earth’s surface and are distributed very heterogeneously \citep{Choulga2014}, analysis of the results here was restricted to areas where there have been significant changes in the surface physiographic fields, i.e. a change in any of the surface field when going from V15 to V20 (and to V15X or V20X) of ≥ 10\% (≥ 0.1 for fractional fields); if lake or vegetation cover changed from 0.1 in V15 to 0.3 in V20 field set this change is classified as significant. Choice of ≥ 10 \% cut-off was adopted as it proofed to be a good trade off between having a sufficient number of grid points to inspect and the strength of the effect of changing the input field. As the cut-off \% increases less points are selected even though with more severe changes, whereas when the cut-off \% decreases more points are selected but it becomes more difficult to disentangle the change in the prediction accuracy from VESPER’s training noise. Alternative cut-off \% were briefly explored, but conclusions of the results remained broadly unchanged. All grid-cells selected for the analysis can be classified according to how the surface fields are updated when going from V15 to V20 (note that categories represent a systematic and consistent update across multiple related fields, and do not include any restrictions on other surface fields apart the ones mentioned): 
\begin{itemize}
	\item \textbf{Lake Updates}. The change in the lake cover \textit{cl} and lake depth \textit{dl} are significant, but the changes in ocean and glacier \textit{glm} fractions are not. This corresponds to grid-cells where lakes have been added or removed. Lake-Ground Updates is a sub-category where additional constraint that the change in the high/low vegetation fractions \textit{cvh/cvl} are not significant is in place. This then corresponds to the exchange of lakes for bare ground, or vice versa.
	\item \textbf{Vegetation Updates.} The change in the high vegetation fraction \textit{cvh} is significant, but the change in lake cover \textit{cl} is not significant. This corresponds to grid-cells where large features like forests and woodlands have been updated, exchanged for bare ground or low vegetation.
	\item \textbf{Glacier Updates}. The change in the glacier cover \textit{glm} is significant. This corresponds to any areas where the fraction of glacier ice has been updated.
\end{itemize}
All four VESPER versions were trained on 2016 data over the entire globe, and then used to predict LST for 2019. To compute the training noise (i.e. the network variance – changes in the LST predictability due to model training) every VESPER’s version was in turn trained four times and each time with a different seed. To assess the changes of LST predictability due to the use of the updated surface physiographic fields instead of V15 field set (default) a simple metric $\delta_{\rm VM}$ following:
\begin{equation} 
	\delta_{\rm VM} = \text{MAE}_{\rm VESPER\_VM} - \text{MAE}_{\rm VESPER\_15} 
\end{equation}

%\begin{align}
%\begin{eqnarray}
%	\delta_{\rm VM} = MAE_{\rm VESPER\_VM} -  MAE_{\rm VESPER\_V15} = MAE \left[LST_{\rm VESPER\_VM}-LST_{\rm Aqua-MODIS}\right] - MAE \left[LST_{\rm VESPER\_V15}-LST_{\rm Aqua-MODIS}\right]
%\end{eqnarray}
%\end{align}
where $M$ represents one of the field set versions V20, V20X or V15X, and $\text{MAE}$ is computed over the whole prediction period of 2019; e.g. V20 describes the difference between MAE of differences between VESPER\_V20 LST predictions and Aqua-MODIS LST and between MAE of differences between VESPER\_V15 LST predictions and Aqua-MODIS LST, negative V20 indicates that VESPER\_V20 LST prediction is more accurate and vice versa.

@@

\subsection{Evaluation of updated lake fields}\label{sec:3}
To understand if there is a way to automatically and rapidly assess accuracy of updated and/or new surface physiography fields and if their use in the atmospheric model increase predictability a neural network regression model VESPER was trained over 2016 data and used to predict LST for 2019. VESPER’s training noise was confirmed to be much smaller (\hl{estimated noise bias over all grid-cells for 2019 at ~31km resolution is 0.04K}) than differences in LST predictions by different VESPER configurations (\hl{degrees K}), so all changes in LST predictability can be attributed to the changes in surface physiographic fields. As a first attempt lake related information is assessed, namely lake cover (and land-sea mask and glacier cover as they are used for lake cover generation) and lake mean depth, that were created from scratch using new up-to-date high-resolution input datasets (see Table \ref{tab:datasources}) for the V20 (and V20X) field set; other surface physiographic fields (see Table \ref{table:definitions}) were regenerated from the same input sources as in initial V15 field set but taking into account that lake related fields were changed. In cases when existing in V15 lake cover water was removed in V20, its place could fill high or low vegetation, glacier or bare ground. To analyse results grid-cells were divided into four categories, and each was investigated in detail (see Table \ref{tab:summary_table} for the results aggregated over the whole globe).

\begin{table*}
	\begin{tabular}{ccccccccc}
		\hline
		\multirow{2}{*}{Category} & \multirow{2}{*}{Number of grid cells} & 	\multicolumn{4}{c}{Training noise of VESPER\_VM, $K$} &\multicolumn{3}{c}{$\delta_{\rm VM}$} \\  
		&&V15  & V15X & V20 & V20X & V15X &V20 & V20X  \\
		\hline 
		Lake&1631&0.009 & 0.008 & 0.007 & - &- &-0.37 & -  \\
		Lake-Ground&546&0.02 & 0.03 & 0.02 & - &- &-0.83 & -  \\
		Vegetation&58&0.06 & 0.10 & 0.13 & - & -&+0.03 & -  \\
		Glacier&1057& 0.07& 0.09 & 0.08 & - &-&-0.23  & -  \\
		\bottomrule
	\end{tabular}
	\caption{Globally averaged differences VM between mean absolute error (MAE) of [VESPER\_VM LST predictions minus Aqua-MODIS LST] and MAE of [VESPER\_V15 LST predictions minus Aqua-MODIS LST] for 2019 at ~31km resolution (where M denotes V15X, V20, V20X field sets) per grid-cell category. Negative VM values indicate increase of LST predictability due to the use of the updated surface physiographic fields instead of V15 field set (default), positive VM values indicate decrease of LST predictability and suggest the presence of erroneous information in the surface physiographic fields. Training noise values are much smaller than the variance between different VESPER’s configurations, indicating that changes in LST predictability are mainly due to changes in the surface physiographic fields. \hl{TK: TBD}}
	\label{tab:categorisation}
\end{table*}

@@

@@

		&&V15  & V15X & V20 & V20X &delta &delta & delta  \\

@@

		\multirow{9}{*}{Lake}&Lake Natron centre, Tanzania&0 & 0 & 0 & 0 &1 &1 & 1  \\
		&Lake Natron north, Tanzania&0 & 0 & 0 & 0 &1 &1 & 1  \\
		&Lake Blanche, Australia&0 & 0 & 0 & 0 &1 &1 & 1  \\
		&Great Salt Lake Desert, Utah&0 & 0 & 0 & 0 &1 &1 & 1  \\
		&Farah Province, Afghanistan&0 & 0 & 0 & 0 &1 &1 & 1  \\
		&Gujarat Province, India&0 & 0 & 0 & 0 &1 &1 & 1  \\
		&Toshka Lakes, Egypt&0 & 0 & 0 & 0 &1 &1 & 1  \\
		&North-West Canada&0 & 0 & 0 & 0 &1 &1 & 1  \\
		&Astrakham Nature reserve, north edge east of the Caspian Sea&0 & 0 & 0 & 0 &1 &1 & 1  \\

@@

		Lake-ground&-&0 & 0 & 0 & 0 &1 &1 & 1  \\

@@

		\multirow{2}{*}{Glacier}&Bering Glacier, Alaskan gulf&0 & 0 & 0 & 0 &1 &1 & 1  \\
		&Juncal Glacier, Chilean Andes &0 & 0 & 0 & 0 &1 &1 & 1  \\

>>>>

\noindent As focus of this study are lake related fields, and lakes occupy only 1.8\% of the Earth’s surface and are distributed very heterogeneously \citep{Choulga2014}, analysis of the results here was restricted to areas where there have been significant changes in the surface physiographic fields, i.e. a change in any of the surface field when going from V15 to V20 (and to V15X or V20X) of ≥ 10\% (≥ 0.1 for fractional fields); if lake or vegetation cover changed from 0.1 in V15 to 0.3 in V20 field set this change is classified as significant. Choice of ≥ 10 \% cut-off was adopted as it proofed to be a good trade off between having a sufficient number of grid points to inspect and the strength of the effect of changing the input field. As the cut-off \% increases less points are selected even though with more severe changes, whereas when the cut-off \% decreases more points are selected but it becomes more difficult to disentangle the change in the prediction accuracy from VESPER’s training noise. Alternative cut-off \% were briefly explored, but conclusions of the results remained broadly unchanged. All grid-cells selected for the analysis can be classified according to how the surface fields are updated when going from V15 to V20 (note that categories represent a systematic and consistent update across multiple related fields, and do not include any restrictions on other surface fields apart the ones mentioned): 
\begin{itemize}
	\item \textbf{Lake Updates}. The change in the lake cover \textit{cl} and lake depth \textit{dl} are significant, but the changes in ocean and glacier \textit{glm} fractions are not. This corresponds to grid-cells where lakes have been added or removed. Lake-Ground Updates is a sub-category where additional constraint that the change in the high/low vegetation fractions \textit{cvh/cvl} are not significant is in place. This then corresponds to the exchange of lakes for bare ground, or vice versa.
	\item \textbf{Vegetation Updates.} The change in the high vegetation fraction \textit{cvh} is significant, but the change in lake cover \textit{cl} is not significant. This corresponds to grid-cells where large features like forests and woodlands have been updated, exchanged for bare ground or low vegetation.
	\item \textbf{Glacier Updates}. The change in the glacier cover \textit{glm} is significant. This corresponds to any areas where the fraction of glacier ice has been updated.
\end{itemize}
The VESPER models is trained on 2016 data over the entire globe, and then used to predict LST for 2019. The training of a neural network is inherently stochastic - the same model trained twice with the same data can settle in different local optima and so make different predictions. To make our conclusions robust against this training noise, each VESPER model is in turn trained 4 times. For each MODIS ground truth we then have 4 LST predictions per model. To assess the changes of LST predictability due to the use of the updated surface physiographic fields instead of V15 field set (default) we compare the mean absolute error (MAE) between different VESPER models using the a simple metric $\delta_{\rm VM}$ is used:
\begin{equation} 
	\delta_{\rm VM} = \text{MAE}_{\rm VESPER\_VM} - \text{MAE}_{\rm VESPER\_V15} 
\end{equation}
where $M$ represents one of the field set versions V20, V20X or V15X, and $\text{MAE}$ is computed over the whole prediction period of 2019. In turn, the MAE is the error between the prediction of a VESPER model and the Aqua-MODIS LST. A negative $\delta_{\rm VM}$ indicates that the VESPER\_VM LST prediction is more accurate than the  VESPER\_V15 prediction, and vice versa.

@@

\subsection{Evaluation of updated lake fields}\label{sec:3}
To understand if there is a way to automatically and rapidly assess accuracy of updated and/or new surface physiography fields and if their use in the atmospheric model increase predictability a neural network regression model VESPER was trained over 2016 data and used to predict LST for 2019.

VESPER’s training noise was confirmed to be much smaller (\hl{estimated noise bias over all grid-cells for 2019 at ~31km resolution is 0.04K}) than differences in LST predictions by different VESPER configurations (\hl{degrees K}), so all changes in LST predictability can be attributed to the changes in surface physiographic fields.

As a first attempt lake related information is assessed, namely lake cover (and land-sea mask and glacier cover as they are used for lake cover generation) and lake mean depth, that were created from scratch using new up-to-date high-resolution input datasets (see Table \ref{tab:datasources}) for the V20 (and V20X) field set; other surface physiographic fields (see Table \ref{table:definitions}) were regenerated from the same input sources as in initial V15 field set but taking into account that lake related fields were changed. In cases when existing in V15 lake cover water was removed in V20, its place could fill high or low vegetation, glacier or bare ground. To analyse results grid-cells were divided into four categories, and each was investigated in detail (see Table \ref{tab:summary_table} for the results aggregated over the whole globe).

\begin{table*}
	\begin{tabular}{ccccccccc}
		\toprule
		\multirow{2}{*}{Category} & \multirow{2}{*}{Number of grid cells} & 	\multicolumn{4}{c}{Training noise of VESPER\_VM, $K$} &\multicolumn{3}{c}{$\delta_{\rm VM}$, $K$} \\  
		&&V15  & V15X & V20 & V20X & V15X &V20 & V20X  \\
		\hline 
		Lake&1631&0.09 & 0.09 & 0.08 & 0.09 &$-0.20$ &$-0.37$ & $-0.37$  \\
		Lake-Ground&546&0.16 & 0.17 & 0.15 & 0.09 &$-0.56$ &$-0.83$ & $-1.12$ \\
		Vegetation&58&0.25 & 0.31 & 0.36& 0.46 & $-4 \times 10^{-3}$&+0.04 & $+3 \times 10^{-4}$  \\
		Glacier&1057& 0.27& 0.30& 0.28 & 0.25 &$-9\times 10^{-3}$&$-0.22$ & $-0.28$  \\
		\bottomrule
	\end{tabular}
	\caption{Globally averaged differences VM between mean absolute error (MAE) of [VESPER\_VM LST predictions minus Aqua-MODIS LST] and MAE of [VESPER\_V15 LST predictions minus Aqua-MODIS LST] for 2019 at ~31km resolution (where M denotes V15X, V20, V20X field sets) per grid-cell category. Negative VM values indicate increase of LST predictability due to the use of the updated surface physiographic fields instead of V15 field set (default), positive VM values indicate decrease of LST predictability and suggest the presence of erroneous information in the surface physiographic fields. Training noise values are much smaller than the variance between different VESPER’s configurations, indicating that changes in LST predictability are mainly due to changes in the surface physiographic fields. \hl{TK: TBD}}
	\label{tab:categorisation}
\end{table*}

\subsubsection{Category: Lake updates}
\begin{figure}
	\includegraphics[width=\columnwidth]{categories_lakes_histogram}
	\caption{some caption} 
	\label{fig:example_figure_histogram}
\end{figure}

@@

\begin{figure}
	\includegraphics[width=\columnwidth]{categories_v15_v20}
	\caption{some caption} 
	\label{fig:lake_sub_categories}
\end{figure}

@@

		&&V15  & V15X & V20 & V20X & V15X &V20 & V20X  \\

@@

		\multirow{9}{*}{Lake}&Lake Natron centre, Tanzania&0.14 & 1.70 & 0.93 & 0.61 &1.35 &2.45 & 2.61  \\
		&Lake Natron north, Tanzania&0.15 & 0.43 & 0.59 & 0.20 &0.72 &1.57 & 1.24  \\
		&Lake Blanche, Australia&0.96 & 0.80 & 1.78 & 0.32 &-0.42 &0.03 & -0.75  \\
		&Great Salt Lake Desert, Utah&0.30 & 0.48 & 1.06 & 0.71 &-0.18 &2.91 & 0.25	  \\
		&Farah Province, Afghanistan&0.016 & 0.22 & 0.10 & 0.25 &0.22 &0.19 & -0.13  \\
		&Gujarat Province, India&2.93 & 1.30 & 0.48 & 1.20 &-1.26 &4.21 & \textcolor{red}{5.24}  \\
		&Toshka Lakes, Egypt&1.05 & 0.46 & 0.30 & 0.16 &-0.39 &\textcolor{red}{-1.14} & -1.28  \\
		&North-West Canada&0.07 & 0.06 & 0.05 & 0.06 &-0.04 &-0.067 & -0.067  \\
		&Astrakham Nature reserve, north edge east of the Caspian Sea&0.20 & 0.30 & 0.26 & 0.30 &0.05 &0.012 & -0.007  \\

@@

		Lake-ground&-& &  &  &  & & &   \\

@@

		\multirow{2}{*}{Glacier}&Bering Glacier, Alaskan gulf&0.59 & 0.56 & 0.25 & 0.090 &-0.54 &-0.16 & -0.80  \\
		&Juncal Glacier, Chilean Andes &0.78 & 0.34 & 1.22 & 0.41 &0.11 &1.26 & 1.20  \\
